In this section we will pause to construct an important family of example rings. 

\begin{dfn}[Quadratic Extension]
Let \(R\) be a commutative ring, and let \(D \in R\). We define the \emph{quadratic extension} \index{quadratic extension} of \(R\) by \(D\) to be \(R[\sqrt{D}] = R \times R\) as a set, and define operations on \(R[\sqrt{D}]\) as follows.
\begin{eqnarray*}
(r_1,r_2) + (s_1,s_2) & = & (r_1+s_1, r_2+s_2) \\
(r_1,r_2) \cdot (s_1,s_2) & = & (r_1s_1 + Dr_2s_2, r_1s_2 + r_2s_1)
\end{eqnarray*}
We will write the element \((a,b) \in R[\sqrt{D}]\) as \(a + b\sqrt{D}\).
\end{dfn}

The notation \(R[\sqrt{D}]\) is intended to suggest that this structure behaves much like \(R\), but now the element \(D\) has a ``square root''. It's important to remember that the plus in \(a + b\sqrt{D}\) is a \emph{structural} symbol, not an operation symbol.

\begin{prop} \label{prop:quad-ext-ring}
Let \(R\) be a ring and \(D \in R\).
\begin{proplist*}
\item \(R[\sqrt{D}]\) is a commutative ring.
\item If \(R\) is unital, then \(R[\sqrt{D}]\) is unital.
\item The map \(\iota : R \rightarrow R[\sqrt{D}]\) given by \(\iota(r) = (r,0_R)\) is a ring homomorphism. If \(R\) is unital, then \(\iota\) is unital.
\end{proplist*}
\end{prop}

The most important use of quadratic extensions is to construct new domains and fields out of old ones.

\begin{prop} \label{prop:quad-ext-dom} \mbox{}
\begin{proplist}
\item If \(R\) is a domain, then \(R[\sqrt{D}]\) is a domain if and only if the equation \(Dy^2 = x^2\) has no nonzero solutions in \(R\).
\item If \(R\) is a field, then \(R[\sqrt{D}]\) is a field if and only if the equation \(D = x^2\) has no nonzero solutions in \(R\).
\end{proplist}
\end{prop}

\begin{proof}
\begin{inlineproplist}
\item (@@@)
\item (@@@)
\end{inlineproplist}
\end{proof}

\begin{prop} \label{prop:sqfree-int}
Let \(D\) be an integer. We say that \(D\) not equal to 0 or 1 is \emph{squarefree} \index{squarefree!integer} if \(D\) is not divisible by \(a^2\) for any \(a > 1\).
\begin{proplist}
\item If \(D\) is squarefree, then the equation \(Dy^2 = x^2\) has no nonzero solutions in \(\ZZ\), and thus \(\ZZ[\sqrt{D}]\) is a domain. \label{prop:sqfree-int:zz-dom}
\item If \(D\) is squarefree, then the equation \(D = x^2\) has no solutions in \(\QQ\), and thus \(\QQ[\sqrt{D}]\) is a field. \label{prop:sqfree-int:qq-field}
\end{proplist}
\end{prop}

\begin{proof}
\begin{inlineproplist}
\item We consider two cases: if \(|D| \leq 1\) and if \(|D| > 1\). In the first case, \(D = -1\). Now if \(Dy^2 = x^2\), then we have \(-y^2 = x^2\); by sign considerations we have \(x = y = 0\). In the second case, \(D\) is divisible by a prime, say \(p\). If \(x = 0\), then \(y = 0\), and likewise if \(y = 0\) then \(x = 0\), so we can assume \(x\) and \(y\) are both nonzero. By the fundamental theorem of arithmetic, \(p\) appears the same number of times in the factorizations of \(Dy^2\) and \(x^2\). But note that \(p\) must appear an even number of times in \(x^2\), but an odd number of times in \(Dy^2\), a contradiction. Thus in either case the equation \(Dy^2 = x^2\) has no nonzero solutions.
\item If \(D = x^2\) has a rational solution \(x = p/q\), then \(Dq^2 = p^2\), which contradicts \ref{prop:sqfree-int:zz-dom}.
\end{inlineproplist}
\end{proof}

There are plenty of squarefree integers; in fact these are precisely the integers with no repeated prime factors. Each such integer yields a different domain \(\ZZ[\sqrt{D}]\) and field \(\QQ[\sqrt{D}]\). These rings are particularly nice for a few reasons. They are easy to compute in, and while similar to \(\ZZ\) and \(\QQ\), can exhibit wildly different behavior, as we will see in the next chapter.




%---------%
\Exercises%
%---------%

\begin{exercise}
Prove Proposition \ref{prop:quad-ext-ring}.
\end{exercise}

\begin{exercise}
Let \(R = \ZZ/(3)\).
\begin{proplist}
\item Show that \(2x^2 = y^2\) has no nonzero solutions in \(R\). (When in doubt, use brute force.)
\item Conclude that \(R[\sqrt{2}]\) is a field with 9 elements.
\end{proplist}
\end{exercise}

\begin{exercise}
Elements of \(\ZZ[\sqrt{-1}]\) are sometimes called \emph{gaussian integers}.
\begin{proplist}
\item Determine which elements of \(\ZZ[\sqrt{-1}]\) are units.
\end{proplist}
\end{exercise}

\begin{exercise}
Show that \(\CHAR{R} = \CHAR{R[\sqrt{D}]}\).
\end{exercise}

\begin{exercise}
Let \(R\) be a ring and \(S \subseteq R\) a subring, with \(D \in S\). Show that \(S[\sqrt{D}]\) is a subring of \(R[\sqrt{D}]\).
\end{exercise}

\begin{exercise}
Let \(\varphi : R \rightarrow S\) be a ring homomorphism with \(D \in R\) and \(E = \varphi(D)\). Show that there is a unique ring homomorphism \(\Phi : R[\sqrt{D}] \rightarrow S[\sqrt{E}]\) such that the following diagram commutes.
\begin{center}
\begin{tikzcd}
R \arrow[r,"\varphi"] \arrow[d, "\iota"']    & S \arrow[d, "\iota"] \\
R[\sqrt{D}] \arrow[r, "\Phi"'] & S[\sqrt{E}]
\end{tikzcd}
\end{center}
\end{exercise}
