\begin{dfn}[Prime] \label{dfn:zz-prime}
We say an integer \(p \notin \{1,0,-1\}\) is \emph{prime} if whenever \(p = ab\), either \(a = \pm p\) or \(b = \pm p\).
Equivalently, \(p\) is prime if it is not 0, 1, or -1, and the only divisors of \(p\) are \(\pm 1\) and \(\pm p\).
\end{dfn}

\begin{prop} \mbox{}
\begin{enumerate}
\item 2 and 3 are prime.
\item If \(p\) is prime, then \(-p\) is prime.
\item If \(p\) and \(q\) are prime such that \(p|q\), then \(q = \pm p\).
\item If \(p\) is a prime integer, then for all integers \(a\), \(\GCD{a}{p}\) is \(|p|\) if \(p|a\) and \(1\) otherwise.
\end{enumerate}
\end{prop}

\begin{prop} \label{prop:zz-prime-2}
An integer \(p\) is prime if and only if whenever \(p|ab\), either \(p|a\) or \(p|b\).
\end{prop}

\begin{proof}
This is an ``if and only if'' statement; we proceed by proving the ``only if'' part and then the ``if'' part.
\begin{inlineproplist}
\item[\((\Rightarrow)\)] Suppose \(p\) is prime and that \(p|ab\).
Consider \(\GCD{a}{p}\).
Since \(p\) is prime, there are two possibilities.
\begin{inlineproplist}
\item If \(\GCD{a}{p} = |p|\), then \(|p|\) divides \(a\), so that \(p|a\).
\item If \(\GCD{a}{p} = 1\), then by Euclid's Lemma, \(p|b\).
\end{inlineproplist}

\item[\((\Leftarrow)\)] Suppose \(p\) has the property that if \(p|ab\) then either \(p|a\) or \(p|b\).
Suppose further that \(p = ab\).
In particular \(p|ab\) (since \(1 \cdot p = ab\)) and so, without loss of generality, \(p|a\).
Say \(a = pa'\).
Now \(p = ab = pa'b\), and thus \(1 = a'b\).
Now \(|b| = 1\), and thus \(|p| = |a|\), so that \(p = \pm a\) as needed.
\end{inlineproplist}
\end{proof}

\begin{cor}
If \(p\) is a prime and \(a_i\) integers such that \(p|a_1a_2 \cdots p_n\), then \(p|a_i\) for some \(i\).
\end{cor}



\begin{thm}[Fundamental Theorem of Arithmetic: Part 1]
Every integer other than \(0\), \(1\), and \(-1\) can be written as a product of primes.
That is, every such \(n\) can be expressed as \(n = p_1p_2 \cdots p_k\), where the \(p_i\) are prime.
This is called a \emph{prime factorization} of \(n\).
\end{thm}

\begin{proof}
Suppose first that \(b > 0\).
We proceed by strong induction.
\begin{proplist}
\item \textbf{Base Case} (\(b = 2\)): If \(d|2\), then \(0 < |d| \leq |2|\).
Thus the only possible divisors of 2 are \(\pm 1\) and \(\pm 2\), and so 2 is prime by definition.
\item \textbf{Inductive Step}: Suppose that for some \(n\), every integer \(2 \leq n' < n\) can be written as a product of primes, and consider \(n\).
If \(n\) is itself prime, then \(n = n\) is its own prime factorization.
If \(n\) is not prime, then by definition there exist integers \(a\) and \(b\) such that \(n = ab\) and \(n \neq \pm a\) and \(n \neq \pm b\).
Since \(n > 0\), we can assume that \(a > 0\) and \(b > 0\).
In fact we have \(a > 1\) and \(b > 1\), so that \(a,b < n\).
By the inductive hypothesis, \(a\) and \(b\) have prime factorizations; say \(a = p_1p_2 \cdots p_h\) and \(b = q_1q_2 \cdots q_k\).
Now \[ n = ab = p_1p_2 \cdots p_h q_1q_2 \cdots q_k \] has a prime factorization.
\end{proplist}
Thus by strong induction every integer \(n \geq 2\) has a prime factorization.
If \(n < 0\), then \(-n > 0\), so that \(-n = p_1p_2 \cdots p_k\) has a prime factorization; then \(n = (-p_1)p_2 \cdots p_k\) also has a prime factorization.
\end{proof}

\begin{thm}[Fundamental Theorem of Arithmetic: Part 2]
The prime factorization of an integer is unique in the following sense.
If \(n\) has two prime factorizations \[ n = p_1p_2 \cdots p_k = q_1q_2 \cdots q_\ell, \] then \(k = \ell\) and, after relabeling the \(q_i\), we have \(p_i = \pm q_i\) for each \(1 \leq i \leq k\).
\end{thm}

\begin{proof}
We saw in FTA Part 1 that every such \(n\) has at least one prime factorization, which consists of at least one prime factor.
To show uniqueness we will proceed by strong induction on the \emph{length} of the shortest prime factorization of \(n\).
\begin{inlineproplist}
\item \textbf{Base Case:} Suppose \(n = p\) has a prime factorization of length 1; that is, \(n\) itself is prime.
Suppose \(p = q_1q_2 \ldots q_\ell\) is another prime factorization of \(n\).
Since \(p\) is prime, we have (rearranging the \(q_i\) if necessary) \(p|q_1\).
Since \(p\) and \(q_1\) are prime, we have \(q_1 = \pm p\).
Now if \(\ell > 1\) we have \[ 1 = |q_2 \cdots q_\ell|, \] so that \(|q_i| = 1\) for each \(q_i\), a contradiction.
So in fact \(\ell = 1\) and \(q_1 = \pm p\), as claimed.
\item \textbf{Inductive Step:} Suppose that every integer having a shortest prime factorization of length at most \(k\) has a unique prime factorization, and suppose \(n = p_1p_2 \cdots p_{k+1}\) is an integer with a shortest prime factorization of length \(k+1\).
Suppose further that \(n = q_1q_2 \cdots q_\ell\) is another prime factorization of \(n\).
In particular, \(p_1 | q_1q_2 \cdots q_\ell\), so that, rearranging the \(q_i\) if necessary, \(p|q_1\).
Now \(q_1 = \pm p\), and we have \[p_2 \cdots p_{k+1} = q_2 \cdots q_\ell. \]
Note that these are two prime factorizations of an integer having a shortest prime factorization of length at most \(k\).
By the Inductive Hypothesis, we have \(\ell = k+1\) and, relabeling the \(q_i\) if necessary, \(q_i = \pm p_i\) for each \(2 \leq i \leq k+1\).
So the prime factorization of \(n\) is unique.
\end{inlineproplist}
\end{proof}

We've established that every integer has an essentially unique prime factorization.
But \textbf{how do we prove that a given integer is prime}?
The definition suggests one way to do it, known as \emph{trial division}: an integer \(n\) is prime if and only if \(n\) is not divisible by any integer \(t\) with \(1 < t < n\).
This works, but is extremely time consuming.
A better method is suggested by the following.

\begin{prop}
Let \(n > 1\) be an integer, and let \(t\) be the largest integer such that \(t^2 < n\).
(Such an integer exists, since the set of all \(t\) with \(t^2 < n\) is bounded above by \(n\).)
(Also, this \(t\) is \(\lfloor \sqrt{n} \rfloor\), but we don't know what \(\sqrt{\cdot}\) means.)
Then \(n\) is prime if and only if \(n\) is not divisible by any prime \(p\) with \(2 \leq p \leq t\).
\end{prop}

\begin{proof}
Certainly if \(n\) is prime it is not divisible by any such \(p\).
We prove the ``only if'' part by contraposition.
Suppose that \(n\) is \emph{not} prime; say \(n = ab\), where \(n \neq a\) and \(n \neq b\).
(We can assume positive signs here since \(n > 0\).)
If \(a\) and \(b\) are both strictly larger than \(t\), then we have \(n > t^2 > ab > n\), a contradiction.
Without loss of generality, then, \(a \leq t\).
In particular, all prime factors of \(a\) are less than \(t\) in absolute value, and so \(n\) has a prime factor \(p\) such that \(2 \leq p \leq t\).
\end{proof}

For example, consider \(n = 5\).
The largest \(t\) such that \(t^2 < 5\) is \(t = 2\), and the only prime \(p\) such that \(2 \leq p \leq 2\) is \(p = 2\).
Since (by the Division Algorithm) we have \(5 = 2 \cdot 2 + 1\), with remainder \(1 \neq 0\), 2 does not divide 5.
So \textbf{5 is prime}.

This result gives us a strategy for finding prime numbers, but it only works if we have a complete list of primes up to \(\sqrt{n}\) to begin with.
In other words, we can find a prime if we start with a list of primes.
This can be made into a reasonably efficient algorithm for finding all the primes up to some bound, called the \emph{Sieve of Eratosthenes}.
There are some interesting questions left unanswered, though.
Given an integer \(n\),
\begin{itemize}
\item ...is \(n\) prime?
\item ...what is the prime factorization of \(n\)?
\item ...how many prime factors does \(n\) have?
\item ...how many \emph{distinct} prime factors does \(n\) have?
\item ...what is the smallest prime factor of \(n\)?
\item ...how difficult is it to answer these questions?
\item ...how difficult is it to \emph{verify} the answers to these questions?
\end{itemize}
