\begin{prop}
Let \(A\), \(O\), and \(B\) be distinct points. A point \(P\) in \(\INTANGLE{A}{O}{B}\) is on the bisector of \(\ANGLE{A}{O}{B}\) if and only if \(\SEGMENT{P}{X} \equiv \SEGMENT{P}{Y}\), where \(X\) is the foot of \(P\) on \(\LINE{O}{A}\) and \(Y\) is the foot of \(P\) on \(\LINE{O}{B}\).
\end{prop}

\begin{proof}
Suppose \(P\) has this property. Now \(\TRIANGLE{O}{P}{X}\) and \(\TRIANGLE{O}{P}{Y}\) are right, with \(\SEGMENT{P}{X} \equiv \SEGMENT{P}{Y}\) and \(\SEGMENT{O}{P} \equiv \SEGMENT{O}{P}\). By the HL Theorem, \(\TRIANGLE{O}{P}{X} \equiv \TRIANGLE{O}{P}{Y}\), and thus \(\ANGLE{X}{O}{P} \equiv \ANGLE{Y}{O}{P}\). So \(P\) is on the bisector of \(\ANGLE{A}{O}{B}\).

Conversely, suppose \(P\) is on the bisector of \(\ANGLE{A}{O}{P}\), and let \(X\) be the foot of \(P\) on \(\LINE{O}{A}\) and \(Y\) the foot of \(P\) on \(\LINE{O}{B}\). Now \(\TRIANGLE{X}{O}{P} \equiv \TRIANGLE{Y}{O}{P}\) by AAS, so that \(\SEGMENT{P}{X} \equiv \SEGMENT{P}{Y}\). 
\end{proof}

\begin{construct}[Incircle Theorem]
Let \(A\), \(B\), and \(C\) be distinct points. Then we have the following.
\begin{enumerate}
\item The bisectors of the interior angles of \(\TRIANGLE{A}{B}{C}\) are concurrent at a point \(O\), called the \emph{incenter} of the triangle.

\item The feet of \(O\) on the sides of \(\TRIANGLE{A}{B}{C}\) lie on a circle, called the \emph{incircle} of \(\TRIANGLE{A}{B}{C}\), which is centered at \(O\) and tangent to the sides of \(\TRIANGLE{A}{B}{C}\).
\end{enumerate}
\end{construct}

\begin{proof}
Let \(\RAY{A}{A'}\) be the bisector of \(\ANGLE{B}{A}{C}\). By the Crossbar Theorem this ray cuts \(\SEGMENT{B}{C}\) at a point \(A''\). Let \(\RAY{B}{B'}\) be the bisector of \(\ANGLE{A}{B}{C}\); again by the Crossbar Theorem this ray cuts \(\SEGMENT{A}{A''}\) at a point \(O\). Let \(X\), \(Y\), and \(Z\) be the feet of \(O\) on \(\LINE{A}{C}\), \(\LINE{A}{B}\), and \(\LINE{B}{C}\), respectively. Since \(O\) is on the bisectors of \(\ANGLE{B}{A}{C}\) and \(\ANGLE{A}{B}{C}\), we have \(\SEGMENT{O}{X} \equiv \SEGMENT{O}{Y}\) and \(\SEGMENT{O}{Y} \equiv \SEGMENT{O}{Z}\); thus \(\SEGMENT{O}{X} \equiv \SEGMENT{O}{Z}\), and so \(O\) is also on the bisector of \(\ANGLE{B}{C}{A}\). Thus the bisectors of the interior angles of \(\TRIANGLE{A}{B}{C}\) are concurrent at \(O\).

Now \(X\), \(Y\), and \(Z\) are the feet of \(O\) on the sides of \(\TRIANGLE{A}{B}{C}\), and we've seen that \(\SEGMENT{O}{X} \equiv \SEGMENT{O}{Y} \equiv \SEGMENT{O}{Z}\). Thus the circle \(\Circle{O}{X}\) contains \(X\), \(Y\), and \(Z\), and moreover is tangent to the sides of \(\TRIANGLE{A}{B}{C}\) at \(X\), \(Y\), and \(Z\).
\end{proof}

\begin{construct}[Excircle Theorem]
Let \(A\), \(B\), and \(C\) be distinct points forming \(\TRIANGLE{A}{B}{C}\). Then we have the following.
\begin{enumerate}
\item The bisector of the interior angle at \(A\) and the exterior angles at \(B\) and \(C\) are concurrent at a point \(O\), called the \emph{excenter} of \(\TRIANGLE{A}{B}{C}\) at \(A\).

\item The feet of \(O\) on the (extended) sides of \(\TRIANGLE{A}{B}{C}\) lie on a circle, called the \emph{excircle} of \(\TRIANGLE{A}{B}{C}\) at \(A\), which is centered at \(O\) and tangent to the sides of \(\TRIANGLE{A}{B}{C}\).
\end{enumerate}
\end{construct}

\begin{proof}
Essentially the same as the proof of the Incircle Theorem.
\end{proof}

To every triangle we can associate four special circles: the incircle, and one excircle for each vertex. These circles are tangent to all three (extended) sides of the circle.

\begin{prop}
Any circle which is tangent to all three (extended) sides of a triangle is either the incircle or one of the excircles.
\end{prop}
