We say that two lines are \emph{perpendicular}\index{perpendicular} if they form a right angle.

\begin{construct}[Foot of a point]\label{construct:foot}
Let \(x\) and \(y\) be distinct points and \(p\) a point not on \(\LINE{x}{y}\).
There is a unique point \(f \in \LINE{x}{y}\) such that \(\LINE{x}{y}\) and \(\LINE{p}{f}\) are perpendicular.
We say that \(f\) is the \emph{foot}\index{foot} of \(p\) on \(\LINE{x}{y}\).
\end{construct}

\begin{proof}
First we show existence.
Note that \(\CIRCLE{x}{p} \cap \CIRCLE{y}{p}\) is not empty (as it contains \(p\)), and by Interleaved Diameters there is a second point \(o\) in the intersection of these circles which is on the opposite side of \(\LINE{x}{y}\).
By the Plane Separation property, \(\LINE{x}{y}\) and \(\SEGMENT{o}{p}\) meet at a unique point \(f\).
Note that \(f\) must be different from either \(x\) or \(y\); WLOG, suppose \(f \neq x\).
Now \(\TRIANGLE{o}{x}{y} \equiv \TRIANGLE{p}{x}{y}\) by SSS, so that \(\ANGLE{p}{x}{f} \equiv \ANGLE{o}{x}{f}\).
Then \(\TRIANGLE{p}{x}{f} \equiv \TRIANGLE{o}{x}{f}\) by SAS.
Then \(\ANGLE{p}{f}{x} \equiv \ANGLE{o}{f}{x}\); that is, \(\ANGLE{p}{f}{x}\) is congruent to its supplement. Thus \(\ell\) and \(\LINE{p}{f}\) are perpendicular.

To see uniqueness, note that if \(p\) has two distinct feet \(f_1\) and \(f_2\) on \(\ell\) then \(p\), \(f_1\), and \(f_2\) form a triangle with two internal right angles, contradicting \ref{cor:a-triangle-has-at-most-one-right-angle}.
\end{proof}

\begin{construct}[Normal at a point]
Let \(\ell\) be a line and \(p \in \ell\) a point in a plane geometry.
There is a unique line \(t\) containing \(p\) which is perpendicular to \(\ell\).
We say that \(t\) is the \emph{normal line}\index{normal line} of \(\ell\) at \(p\).
\end{construct}

\begin{proof}
Let \(x\) be a point on \(\ell\) different from \(p\), and copy \(\SEGMENT{p}{x}\) to the opposite side of \(p\) at a point \(y\) by Circle Separation.
Note that \(p\) is the midpoint of \(\SEGMENT{x}{y}\).
Construct a point \(z\) such that \(\TRIANGLE{x}{y}{z}\) is equilateral.
Now \(\TRIANGLE{z}{x}{p} \equiv \TRIANGLE{z}{y}{p}\) by SSS, so that \(\ANGLE{z}{p}{x} \equiv \ANGLE{z}{p}{y}\), and thus \(\LINE{p}{z}\) is perpendicular to \(\ell\).

Uniqueness follows from the uniqueness of angles on a half-plane, since all right angles are congruent.
\end{proof}

\begin{dfn}[Perpendicular Bisector]
If \(x\) and \(y\) are two points, then the (unique) line perpendicular to \(\LINE{x}{y}\) at the midpoint of \(\SEGMENT{x}{y}\) is called the \emph{perpendicular bisector}\index{perpendicular bisector} of \(\SEGMENT{x}{y}\).
\end{dfn}

\begin{prop}
Let \(x\) and \(y\) be distinct points.
Then \(z\) is on the perpendicular bisector of \(x\) and \(y\) if and only if \(\SEGMENT{x}{z} \equiv \SEGMENT{y}{z}\).
\end{prop}

\subsection*{Intersections of Lines and Circles}

\begin{prop}
In a plane geometry, a line and a circle can have at most two points in common.
\end{prop}

\begin{proof}
Let \(\ell\) be a line and \(\CIRCLE{o}{a}\) a circle which have at least three points in common; say \(x\), \(y\), and \(z\).
Suppose WLOG that \(\BETWEEN{x}{y}{z}\).
Note that \(o\) cannot also be on \(\ell\), as in this case \(z\) cannot be distinct from both \(x\) and \(y\) by the uniqueness of congruent segments on rays.
Now \(\ANGLE{o}{y}{x} \equiv \ANGLE{o}{x}{y}\), \(\ANGLE{o}{y}{z} \equiv \ANGLE{o}{z}{y}\), and \(\ANGLE{o}{x}{z} \equiv \ANGLE{o}{z}{x}\) by Pons Asinorum.
In particular, \(\ANGLE{o}{y}{x}\) is right, so that \(\TRIANGLE{o}{x}{y}\) has two right interior angles -- a contradiction.
\end{proof}

\begin{dfn}[Tangent, Secant]
Let \(\ell\) be a line and \(C\) a circle in a plane geometry.
We say that \(\ell\) is \emph{tangent to} \(C\) if \(\ell\) and \(C\) have exactly one point in common.
Suppose this point is \(t\); in this case we say that \(\ell\) is tangent to \(C\) \emph{at} \(t\).
Similarly, we say that \(\ell\) is a \emph{secant} of \(C\) if \(\ell\) and \(C\) have exactly two distinct points in common.
\end{dfn}

\begin{prop}
Let \(\ell\) be a line and \(C\) a circle with center \(o\) in a plane geometry.
Then \(\ell\) is tangent to \(C\) if and only if \(o\) is not on \(\ell\) and the foot of \(o\) on \(\ell\) is on \(C\).
\end{prop}

\begin{proof}
Suppose \(\ell\) is tangent to \(C\) at \(p\).
If \(o \in \ell\), then \(\ell \cap C\) contains a second point by Circle Separation; so in fact \(o\) is not on \(\ell\).
Let \(f\) be the foot of \(o\) on \(\ell\).
If \(f \neq p\), then \(o\), \(f\), and \(p\) are noncollinear and form a triangle.
Now construct the antipode of \(p\) through \(f\) at \(q\).
Note that \(\TRIANGLE{o}{p}{f} \equiv \TRIANGLE{o}{q}{f}\) by SAS, so that \(\SEGMENT{o}{p} \equiv \SEGMENT{o}{q}\).
In particular, \(q\) is a second point in the intersection of \(\ell\) and \(C\), a contradiction.
Thus \(f = p\) as needed.

Conversely, suppose \(\ell\) does not contain \(o\) and that the foot \(f\) of \(o\) on \(\ell\) is on \(C\).
Suppose BWOC that there is a second point \(g \in \ell \cap C\).
Now \(o\), \(f\), and \(g\) are noncollinear, and \(\SEGMENT{o}{f} \equiv \SEGMENT{o}{g}\), and \(\ANGLE{o}{f}{g}\) is right (by the definition of foot).
So \(\ANGLE{o}{g}{f}\) is right by Pons Asinorum, a contradiction since no triangle can have two interior angles which are both right.
So \(C \cap \ell\) contains exactly one point as needed.
\end{proof}

\begin{construct}[Tangent at a point]
Let \(C\) be a circle with center \(o\) and let \(p\) be a point on \(C\).
There exists a line \(\ell\) which is tangent to \(C\) at \(p\).
\end{construct}

\begin{proof}
Construct the line \(\ell\) which is perpendicular to \(\LINE{o}{p}\) at \(p\).
Then \(o\) is not on \(\ell\), and \(p\) is the foot of \(o\) on \(\ell\).
So \(\ell\) is tangent to \(C\) at \(p\).
\end{proof}

\begin{construct}[Second cut of line and circle]
Let \(\ell\) be a line and \(C\) a circle with center \(o\) in a plane geometry such that \(\ell\) is not tangent to \(C\).
Suppose \(p \in \ell \cap C\).
We may construct the second point in \(\ell \cap C\).
\end{construct}

\begin{proof}
If \(o\) is on \(\ell\), use Circle Separation.
If \(o\) not on \(\ell\), construct the foot \(f\) of \(o\) on \(\ell\).
Using Circle Separation, copy \(\SEGMENT{f}{p}\) onto the opposite side of \(f\) from \(p\) at the point \(q\).
Note that \(\TRIANGLE{o}{f}{p} \equiv \TRIANGLE{o}{f}{q}\) by SAS, so that \(\SEGMENT{o}{p} \equiv \SEGMENT{o}{q}\); thus \(q \in \ell \cap C\) as needed.
\end{proof}

\subsection*{Comparing Segments}

\begin{dfn}
Let \(\SEGMENT{a}{b}\) and \(\SEGMENT{c}{d}\) be segments in a plane geometry.
We say that \(\SEGMENT{a}{b} \leq \SEGMENT{c}{d}\) if there is a point \(x \in \SEGMENT{c}{d}\) such that \(\SEGMENT{a}{b} \equiv \SEGMENT{c}{x}\).
\end{dfn}

\begin{prop} \mbox{}
\begin{enumerate}
\item If \(\SEGMENT{a_1}{b_1} \equiv \SEGMENT{a_2}{b_2}\), \(\SEGMENT{c_1}{d_1} \equiv \SEGMENT{c_2}{d_2}\), and \(\SEGMENT{a_1}{b_1} \leq \SEGMENT{c_1}{d_1}\), then \(\SEGMENT{a_2}{b_2} \leq \SEGMENT{c_2}{d_2}\).
\item If \(\SEGMENT{a}{b} \leq \SEGMENT{c}{d}\) and \(\SEGMENT{c}{d} \leq \SEGMENT{e}{f}\), then \(\SEGMENT{a}{b} \leq \SEGMENT{e}{f}\).
\item If \(\BETWEEN{a}{b}{c}\), then \(\SEGMENT{a}{b} \leq \SEGMENT{a}{c}\).
If \([abcd]\), then \(\SEGMENT{b}{c} \leq \SEGMENT{a}{d}\).
\item If \(\SEGMENT{a}{b} \leq \SEGMENT{c}{d}\) and \(\SEGMENT{c}{d} \leq \SEGMENT{a}{b}\), then \(\SEGMENT{a}{b} \equiv \SEGMENT{c}{d}\).
\end{enumerate}
\end{prop}

\begin{proof}
\item There is a point \(x \in \SEGMENT{c_1}{d_1}\) such that \(\SEGMENT{a_1}{b_1} \equiv \SEGMENT{c_1}{x}\).
Now copy \(\SEGMENT{c_1}{x}\) onto \(\RAY{c_2}{d_2}\) at the point \(y\); note that \(\BETWEEN{c_2}{y}{d_2}\), so that \(y \in \SEGMENT{c_2}{d_2}\).
Now \(\SEGMENT{a_2}{b_2} \equiv \SEGMENT{c_2}{y}\) as needed.

\item There exists a point \(x \in \SEGMENT{c}{d}\) such that \(\SEGMENT{a}{b} \equiv \SEGMENT{c}{x}\), and a point \(y \in \SEGMENT{e}{f}\) such that \(\SEGMENT{c}{d} \equiv \SEGMENT{e}{y}\).
Now copy \(\SEGMENT{c}{x}\) onto \(\RAY{e}{y}\) at the point \(z\); note that \(\BETWEEN{e}{z}{y}\); in particular, \(\SEGMENT{a}{b} \equiv \SEGMENT{e}{z}\).

\item Clear.

\item There exists a point \(x \in \SEGMENT{c}{d}\) such that \(\SEGMENT{c}{x} \equiv \SEGMENT{a}{b}\).
Now either \(x = c\), \(x = d\), or \(\BETWEEN{c}{x}{d}\).
If \(x = c\), then \(b = a\), and \(d = c\), so that \(\SEGMENT{a}{b} \equiv \SEGMENT{c}{d}\).
Suppose \(\BETWEEN{c}{x}{d}\).
There is a point \(y \in \SEGMENT{a}{b}\) such that \(\SEGMENT{c}{y} \equiv \SEGMENT{a}{b}\); but now \(\BETWEEN{a}{b}{y}\), a contradiction.
So we have \(x = d\) as needed.
\end{proof}
